The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-09-23 paper.qmd
editted for susan comment 10

The first method is what we call the "vague question" approach. 
The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how. 
The authors then go on to compare the readers answer to a very specific ground truth. 
This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. 
Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016]. 
For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right". 
The mean of the two groups was used to decide which group was actually "furthest to the right" and the different graphical approaches were evaluated against that ground truth. 
<!-- Participants weren't evaluated against that ground truth, though - the plots were what was actually being evaluated in that study, not participant accuracy. Lineup prompts are typically vague (sometimes, I agree, too vauge) because the participant has to figure out what metric to use to evaluate the data first, and then complete the evaluation task. We're as interested in the metric they're using as we are in their successful identification of the target plot. -->
<!-- Good point about the plots being evaluated, not the participants. I changed that.
I agree with the point about lineups being useful in identifying how participants complete evaluation tasks, and I did comment on that with respect to this particular paper later in the “Qualitative studies” section with this comment:
   “@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper. “
Do you think I should move that comment to this section to keep all comments on that paper together, leave it where it is, or do you think there should be a different edit? -->
In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. 
This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
These studies frequently expect participants to give deterministic answers for probabilistic questions. 
For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". 
The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
<!-- Might be worth referencing the fact that visual statistics and the computational equivalents often differ, and this is even before you start talking measurement error when using a Likert scale.  -->
In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". 
The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. 
<!-- It's not unreasonable to use forms of display that are used in gov't communications, even if they're admittedly not optimal -- it can be reasonable to *show* that they're not optimal, and Padilla is usually advocating for different risk displays as a result, including spaghetti maps for the track and probabilistic wind speed/flood risk maps for the impacts. The problem is that gulf coast residents are used to the bad version and not used to interpreting the other versions, so there's a ton of resistance... and I say this as someone who grew up plotting hurricane paths on maps we got from the grocery store in May at the beginning of hurricane season :). These displays are also complicated by the asymmetry of wind/flood risk around the eye of the storm - the northwest side gets hit harder in the northern hemisphere, which leads to natives referring to the "dirty side" and the "clean side" of the storm. Land topology also shapes flood risk, which is yet another complexity. All of this is to say that the explicitly calculated risk maps do a lot of external cognition, but when you're used to doing that calculus mentally, it can be a bit hard to trust the predigested information. -->
Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
Interestingly, it is unclear how the participants were supposed to answer this question. 
@Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. 
In the first study, participants were required assume to the storm was equally intense regardless of the  probability of the oil rig being hit (i.e. more likely to be hit *does not* mean more damage), but in the second study required participants to assume the opposite (i.e. more likely to be hit *does* mean more damage). 
Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. 
Additionally, by having a ground truth, they end up implicitly asking users to treat uncertainty as a signal which goes against the purposes of signal-suppression.

>>>>

The first method is what we call the "vague question" approach. 
The authors of these studies will ask the participants a question that implies that they should use the uncertainty information, but it is unclear how. 
The authors then go on to compare the readers answer to a very specific ground truth. 
This means the participants are being *evaluated* as though they are performing a value extraction task, but they are not being asked the *questions* that are asked in the value extraction task. 
Ultimately this approach results in strangely cryptic questions that create a large amount of noise due to the varied interpretation of the questions [@Hullman2016].

For example @Hofmann2012 showed study participants 20 plots, where each plot displayed two distributions (as a pair of jittered samples, density plots, histograms, or box plots) and asked them to identify the plot where "the blue group is furthest to the right". 
The mean of the two groups was used to decide which group was actually "furthest to the right" and the different graphical approaches were evaluated against that ground truth. 
<!-- Participants weren't evaluated against that ground truth, though - the plots were what was actually being evaluated in that study, not participant accuracy. Lineup prompts are typically vague (sometimes, I agree, too vauge) because the participant has to figure out what metric to use to evaluate the data first, and then complete the evaluation task. We're as interested in the metric they're using as we are in their successful identification of the target plot. -->
<!-- Good point about the plots being evaluated, not the participants. I changed that.
I agree with the point about lineups being useful in identifying how participants complete evaluation tasks, and I did comment on that with respect to this particular paper later in the “Qualitative studies” section with this comment:
   “@Hofmann2012 tried to capture this by asking participants why they considered a particular plot to be more "right shifted", even though this qualitative analysis did not make it into the final paper. “
Do you think I should move that comment to this section to keep all comments on that paper together, leave it where it is, or do you think there should be a different edit? -->
In another example @Ibrekk1987 asked participants for the "best estimate", but the participant's responses were evaluated against the mean of the distribution.
Ultimately, the term "best" is up to the users interpretation, and the estimate that minimized the sum of squared errors was not implied by the question. 
This vague question approach leads to inconclusive results, as we are left unclear if it was the phrasing of the question or the plot design that caused the participants to answer incorrectly.

A variation of the "vague question" problem, is a series of studies that ask questions that are impossible to answer using the information given to the participants. 
These studies frequently expect participants to give deterministic answers for probabilistic questions. 
For example @Correll2014 showed participants the distribution of voter preferences for two candidates in an election, and asked them "how likely is candidate B to win the election?". 
Participants were not able to answer the question about likelihood in term of probability, but were instead given seven options from 1 = "Outcome will be most in favour of A" to 7 = "Outcome will be most in favour of B". 
The ground truth statistic for this question was a scalar multiple of Cohen’s d, indicating participants were supposed to incorporate uncertainty information using a very specific formula that was likely unknown to them but assumed to be used implicitly.  
In another example, @Padilla2017 provided participants with a visualisation of the cone of uncertainty and asked then to "decide which oil rig will receive more damage based on the depicted forecast of the hurricane path". 
The cone of uncertainty provides a 60% confidence interval for the location of the eye of a hurricane, which allows us to know the area where the eye of the storm will go, but it does not given any information about the intensity of a storm, the size of a storm, or even if a location will be hit. 
<!-- It's not unreasonable to use forms of display that are used in gov't communications, even if they're admittedly not optimal -- it can be reasonable to *show* that they're not optimal, and Padilla is usually advocating for different risk displays as a result, including spaghetti maps for the track and probabilistic wind speed/flood risk maps for the impacts. The problem is that gulf coast residents are used to the bad version and not used to interpreting the other versions, so there's a ton of resistance... and I say this as someone who grew up plotting hurricane paths on maps we got from the grocery store in May at the beginning of hurricane season :). These displays are also complicated by the asymmetry of wind/flood risk around the eye of the storm - the northwest side gets hit harder in the northern hemisphere, which leads to natives referring to the "dirty side" and the "clean side" of the storm. Land topology also shapes flood risk, which is yet another complexity. All of this is to say that the explicitly calculated risk maps do a lot of external cognition, but when you're used to doing that calculus mentally, it can be a bit hard to trust the predigested information. -->
Other authors have commented on the complexity of communicating hurricane risk because the path, storm surge and wind speed are all important and cannot be ignored [@Spiegelhalter2017]. 
Interestingly, it is unclear how the participants were supposed to answer this question. 
@Padilla2017 did three experiments and the first and third experiment had conflicting assumptions in their ground truth. 
In the first study, participants were required assume to the storm was equally intense regardless of the  probability of the oil rig being hit (i.e. more likely to be hit *does not* mean more damage), but in the second study required participants to assume the opposite (i.e. more likely to be hit *does* mean more damage). 
Similar to the vague question problem, these studies seem to be aware that we should be evaluating an uncertainty visualization based on a deterministic observation (such as our example of "does this map have a spatial trend") but are unsure how to incorporate or evaluate it.

While these approaches are certainly a step in the right direction, the experiments end up having far too much noise in their results. 
This is further complicated by the fact that visual statistics and their computational equivalents often differ, even before we consider these measurement errors.
Additionally, by having a ground truth, they end up implicitly asking users to treat uncertainty as a signal which goes against the purposes of signal-suppression.
